The benefits of reporting critical effect size values

### 2024-08-14 paper/paper.Rmd
added abstract and keywords

abstract: |
  Scrivere abstract
keywords          : "keywords"
wordcount         : "X"

>>>>

abstract: |
  Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty on the expected effect size. To assist researchers in calculating critical values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.

keywords          : critical effect size values, statistical significance, fucntions implementation 
wordcount         : "X"

### 2024-08-14 paper/paper.tex
added abstract and keywords

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Scrivere abstract
}

@@

To provide more tangible illustrations of the practical application of critical values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87, p (one-tailed) < .05\). Based on the reported data, Cohen's \(d = 0.736\), indicating a considerable effect size. However, the critical value for Cohen's d in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689'003\), the researchers observed a significant effect worsened emotional states when positive posts were reduced, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results of a similar, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's d in this case are \(-0.006\) and \(+0.006\), emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual \(d = 0.02\)) is truly meaningful in individuals' lives.

@@

In this section, we provide guidance and formulas for computing critical values in general, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's d), correlations (Pearson's r), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package ``criticalESvalues'' accessible at: {[}xxxxx{]}, and are elaborated upon in the subsequent section.

@@

The output now gives a wider range of values: the Cohen's d calculated on the data (\(d\)), the critical Cohen's d (\(dc\)), the numerator of the formula for the critical Cohen's d (\(bc\)), the Cohen's d adjusted for small samples (\(g\)) and the critical Cohen's d adjusted for small samples (\(gc\)).

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical significance values beyond Cohen's d and correlation to include linear regression with both raw and standardized coefficients. This serves as a first step in computing critical values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{critical effect size values, statistical significance, fucntions implementation\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The benefits of reporting critical effect size values},
  pdfauthor={Ambra Perugini1, Filippo Gambarota1, Enrico Toffalini2, Daniël Lakens3, Massimiliano Pastore1, Livio Finos4, Psicostat1, \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={critical effect size values, statistical significance, fucntions implementation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty on the expected effect size. To assist researchers in calculating critical values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.
}

@@

To provide more tangible illustrations of the practical application of critical values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87, p (one-tailed) < .05\). Based on the reported data, Cohen's \(d = 0.736\), indicating a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689'003\), the researchers observed a significant effect worsened emotional states when positive posts were reduced, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results of a similar, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's \(d\) in this case are \(-0.006\) and \(+0.006\), emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual \(d = 0.02\)) is truly meaningful in individuals' lives.

@@

In this section, we provide guidance and formulas for computing critical values in general, with examples for frequently encountered effect sizes including Standardized Mean Differences (Cohen's \(d\)), correlations (Pearson's r), and raw and standardized coefficients in linear models. These formulas have been incorporated into R functions of the package ``criticalESvalues'' accessible at: {[}xxxxx{]}, and are elaborated upon in the subsequent section.

@@

The output now gives a wider range of values: the Cohen's \(d\) calculated on the data (\(d\)), the critical Cohen's \(d\) (\(dc\)), the numerator of the formula for the critical Cohen's \(d\) (\(bc\)), the Cohen's \(d\) adjusted for small samples (\(g\)) and the critical Cohen's \(d\) adjusted for small samples (\(gc\)).

@@

Real-case scenarios may not always be that simple. Hence, we chose to expand the application of computing critical significance values beyond Cohen's \(d\) and correlation to include linear regression with both raw and standardized coefficients. This serves as a first step in computing critical values for a wider array of effects encountered in practical scenarios, where linear models and their extensions are commonly utilized for modeling purposes. A prerequisite is that researchers must be able to identify what parameters in their statistical models reflect the effect sizes of interest, and that they are able to assess their relevance. Notably, however, this prerequisite aligns with the requirements of APA style guidelines concerning the reporting of effect sizes. For further illustration and application, additional examples are provided in the Supplementary online material.
